Certified Symbolic Finite Transducers: Formalization and Applications to String Analysis

### 2025-09-11 Paper/application.tex
polish introduction

Figure \ref{fig:html-sanitizer-sft} illustrates the SFT representation of the HTML encoding sanitizer function.
 The set of safe characters is defined as $\texttt{safe\_chars} = [\texttt{' '}, \texttt{','}, \texttt{'.'}, \texttt{'-'}, \texttt{'\_'}] \cup [\texttt{'0'}-\texttt{'9'}] \cup [\texttt{'A'}-\texttt{'Z'}] \cup [\texttt{'a'}-\texttt{'z'}]$, which can be efficiently represented using the interval algebra we formalized. Unsafe characters are transformed using the encoding function $\texttt{encode}(c) = \texttt{``\&\#''} + \texttt{toString}(\texttt{ord}(c)) + \texttt{``\;''}$. When the alphabet is set to Unicode $[\mathtt{0x0000}, \mathtt{0x10FFFF}]$, the complement $\neg\texttt{safe\_chars}$ represents all Unicode characters outside the safe set, which can also be efficiently represented using our formalized interval operations. \texttt{id}, \texttt{encode}, and $\lambda \_.\varepsilon$ are the identity, encode of unsafe characters, and empty word output functions, respectively.
\begin{figure}[htbp]
\centering
\begin{tikzpicture}[shorten >=1pt, node distance=2.5cm, on grid, auto]
  % States
  \node[state, initial, accepting] (q0) {$q_0$};
  \node[state] (q1) [right=3.5cm of q0] {$q_1$};
  \node[state] (q2) [below=of q0] {$q_2$};

@@

Beyond performance evaluation, our SFT formalization also supports the formal verification of string sanitizer correctness. A sanitizer transforms user inputs into safe outputs, ensuring that the result contains no unsafe characters. However, sanitizers—like any other programs—are error-prone. Manual reviews or testing alone are often insufficient to guarantee their correctness.

@@

We evaluate a range of attack models for HTML, JavaScript, JSON, XML, and CSS. Listing \ref{lst-css-attack} shows the CSS attack model, which is used to verify the corresponding sanitizer. The attack model includes comments, declaration breakers, and known dangerous tokens such as \textbf{expression(}, \textbf{@import}, \textbf{behavior:}, \textbf{-moz-binding}, and \textbf{url(} (unless the URL is separately validated). Sanitizers include GA HtmlEscape, GA JSONEscape, GA XMLEscape, escapeString, and CleanseCSS, where "GA" means from Google Analytics. Sanitizers are often composed with preprocessing steps such as Trim, which removes leading and trailing whitespace characters. As shown in the table, GAHtmlEscape and escapeString are unsafe with respect to the HTML attack model without preprocessing, but become safe when combined with Trim. The \textbf{Size} column demonstrates the expressive power and succinctness of SFTs in modeling real-world transducers. We did not list the execution time of verification because all of them are quite efficent with less than 0.02 second to finish the verificaiton.

>>>>

Figure \ref{fig:html-sanitizer-sft} illustrates the SFT representation of the HTML encoding sanitizer function.
 The set of safe characters is defined as $\texttt{safe\_chars} = [\texttt{' '}, \texttt{','}, \texttt{'.'}, \texttt{'-'}, \texttt{'\_'}] \cup [\texttt{'0'}-\texttt{'9'}] \cup [\texttt{'A'}-\texttt{'Z'}] \cup [\texttt{'a'}-\texttt{'z'}]$, which can be efficiently represented using the interval algebra we formalized. Unsafe characters are transformed using the encoding function $\texttt{encode}(c) = \texttt{``\&\#''} + \texttt{toString}(\texttt{ord}(c)) + \texttt{``;''}$. When the alphabet is set to Unicode $[\mathtt{0x0000}, \mathtt{0x10FFFF}]$, the complement $\neg\texttt{safe\_chars}$ represents all Unicode characters outside the safe set, which can also be efficiently represented using our formalized interval operations. \texttt{id}, \texttt{encode}, and $\lambda \_.\varepsilon$ are the identity, encode of unsafe characters, and empty word output functions, respectively.
\begin{figure}[htbp]
\centering
\begin{tikzpicture}[shorten >=1pt, node distance=2.5cm, on grid, auto]
  % States
  \node[state, initial, accepting] (q0) {$q_0$};
  \node[state] (q1) [right=3.5cm of q0] {$q_1$};
  \node[state] (q2) [below=of q0] {$q_2$};

@@

Beyond performance evaluation, our SFT formalization also supports the formal verification of string sanitizer's safety property. A sanitizer transforms user inputs into safe outputs, ensuring that the result contains no unsafe characters. However, sanitizers—like any other programs—are error-prone. Manual reviews or testing alone are often insufficient to guarantee their correctness.

@@

We evaluate a range of attack models for HTML, JavaScript, JSON, XML, and CSS. Listing \ref{lst-css-attack} shows the CSS attack model, which is used to verify the corresponding sanitizer. The attack model includes comments, declaration breakers, and known dangerous tokens such as \textbf{expression(}, \textbf{@import}, \textbf{behavior:}, \textbf{-moz-binding}, and \textbf{url(} (unless the URL is separately validated).
These are unsafe and undesired strings in CSS contexts.
Sanitizers include GA HtmlEscape, GA JSONEscape, GA XMLEscape, escapeString, and CleanseCSS, where "GA" means from Google Analytics. Sanitizers are often composed with preprocessing steps such as Trim, which removes leading and trailing whitespace characters. As shown in the table, GAHtmlEscape and escapeString are unsafe with respect to the HTML attack model without preprocessing, but become safe when combined with Trim. The \textbf{Size} column demonstrates the expressive power and succinctness of SFTs in modeling real-world transducers. We did not list the execution time of verification because all of them are quite efficent with less than 0.02 second to finish the verificaiton.

### 2025-09-11 Paper/formalization.tex
polish introduction

\paragraph{The product operation formalization.} We formalize the product operation between an SFT $\mathcal{T}$ and an SFA $\mathcal{A}$, denoted as $\mathcal{T} \times \mathcal{A}$ in Section \ref{sec:formalization}. The SFA is formalized as a record type \texttt{NFA in Figure \ref{fig-def-FAs}}.
An additional consideration in this operation is the presence of $\varepsilon$-transitions in our SFT formalization, which implies that the resulting automata may also contain $\varepsilon$-transitions.
%
Therefore, we need to formalize $\varepsilon$SFAs, which is shown in Fig. \ref{fig-def-FAs} as \texttt{eNFA}. The $\varepsilon$SFA formalization extends the standard SFA structure by introducing $\Delta_e'$, which captures $\varepsilon$-transitions as pairs of states, while maintaining the same labeled transition relation $\Delta$ as in standard SFAs.
In this paper, we present only the definitions of  $\varepsilon$SFAs and SFAs, while the complete formalization, including correctness proofs, is available in our Isabelle development.

@@

\begin{figure}[hbt!]
\begin{lstlisting}[mathescape=true]
locale bool_algebra =
 fixes sem :: " 'b $\Rightarrow$ 'a::ord set"
 fixes empty:: "'b $\Rightarrow$ bool"
 fixes nempty:: "'b $\Rightarrow$ bool"
 fixes intersect:: "'b $\Rightarrow$ 'b $\Rightarrow$ 'b"
 fixes diff:: "'b $\Rightarrow$ 'b $\Rightarrow$ 'b"
 fixes elem:: "'a  $\Rightarrow$ 'b $\Rightarrow$ bool"
 assumes
  empty_sem: "empty s = (sem s = {})"
                                    and
  nempty_sem: "nempty_intervals s =
                     ($\neg$ (empty s))" and
  inter_sem: "sem (intersect s1 s2) =
               (sem s1) $\cap$ (sem s2)" and
  diff_sem: "sem (diff f1 f2 s1 s2) =
               (sem s1) - (sem s2)" and
  elem_sem: "elem a s $\equiv$  (a $\in$ sem s)"
\end{lstlisting}
\caption{The boolean algebra locale}
\label{fig-bool-algebra-locale}
\end{figure}

@@

The implementation of \texttt{productT\_imp} follows \emph{Refine\_Monadic} framework's interfaces for sets to store states and transition relations. 
The \emph{Refine\_Monadic} framework provides a way to automatically refine these interfaces to more efficient data structure, such as red-black trees or hashmaps. 
In our formalization, we refine the sets of storing states and transitions to red-black trees.

>>>>

\paragraph{The product operation formalization.} We formalize the product operation between an SFT $\mathcal{T}$ and an SFA $\mathcal{A}$, denoted as $\mathcal{T} \times \mathcal{A}$ in Section \ref{sec:sft}. The SFA is formalized as a record type \texttt{NFA in Figure \ref{fig-def-FAs}}.
An additional consideration in this operation is the presence of $\varepsilon$-transitions in our SFT formalization, which implies that the resulting automata may also contain $\varepsilon$-transitions.
%
Therefore, we need to formalize $\varepsilon$SFAs, which is shown in Fig. \ref{fig-def-FAs} as \texttt{eNFA}. The $\varepsilon$SFA formalization extends the standard SFA structure by introducing $\Delta_e'$, which captures $\varepsilon$-transitions as pairs of states, while maintaining the same labeled transition relation $\Delta$ as in standard SFAs.
In this paper, we present only the definitions of  $\varepsilon$SFAs and SFAs, while the complete formalization, including correctness proofs, is available in our Isabelle development.

@@

\begin{figure}[hbt!]
\begin{lstlisting}[mathescape=true]
locale bool_algebra =
 fixes sem :: " 'b $\Rightarrow$ 'a::ord set"
 fixes empty:: "'b $\Rightarrow$ bool"
 fixes nempty:: "'b $\Rightarrow$ bool"
 fixes intersect:: "'b $\Rightarrow$ 'b $\Rightarrow$ 'b"
 fixes diff:: "'b $\Rightarrow$ 'b $\Rightarrow$ 'b"
 fixes elem:: "'a  $\Rightarrow$ 'b $\Rightarrow$ bool"
 assumes
  empty_sem: "empty s = (sem s = {})"
                                    and
  nempty_sem: "nempty s = $\neg$ (empty s)" 
                                    and
  inter_sem: "sem (intersect s1 s2) =
               (sem s1) $\cap$ (sem s2)" and
  diff_sem: "sem (diff f1 f2 s1 s2) =
               (sem s1) - (sem s2)" and
  elem_sem: "elem a s $\equiv$  (a $\in$ sem s)"
\end{lstlisting}
\caption{The boolean algebra locale}
\label{fig-bool-algebra-locale}
\end{figure}

@@

The implementation of \texttt{productT\_imp} takes advantages of \emph{Refine\_Monadic} framework's interfaces for sets to store states and transition relations. 
The \emph{Refine\_Monadic} framework provides a way to automatically refine these interfaces to more efficient data structure, such as red-black trees or hashmaps. 
In our formalization, we refine the sets of storing states and transitions to red-black trees.

### 2025-09-11 Paper/introduction.tex
polish introduction

Symbolic Finite Automata (SFA) and Symbolic Finite Transducers (SFT)
\cite{cav/DAntoniV17, VeanesHLMB12Transducer,DV21} were proposed around 15 years ago
to deal with the alphabet-explosion problem that often arises in practical
applications. 
%represent advanced extensions of classic FAs and  FTs, enhancing their applicability in practical scenarios.
%
These symbolic extensions permit transition labels defined by an effective 
boolean algebra, allowing for a succinct representation of a set of transitions
that can be conveniently manipulated. 
%more expressive representations. 
For example, in the interval algebra, a transition label might be specified as 
an interval ($'\texttt{a}'-'\texttt{z}'$), encompassing all characters from
$'\texttt{a}'$ to $'\texttt{z}'$. As another example, in the SMT-Algebra
(particularly, of the theory of Linear Integer Arithmetic), %or 
one could associate an arithmetic condition ($x \mod 2 = 0$) to a transition, 
denoting all even numbers. Efficient algorithms for manipulating (e.g. computing
closures) and analyzing (e.g. checking emptiness) SFA and SFT are by now
rather mature (e.g see the CACM article \cite{DV21}).
SFA and SFT have been applied in numerous domains including analysis of web
applications against XSS vulnerabilities \cite{VeanesHLMB12Transducer,uss/HooimeijerLMSV11},
runtime behavior monitoring \cite{osdi/YaseenABCL20}, automated program
transformation \cite{pldi/HuD17}, and finally SMT solving for the unicode theory
of strings (e.g. see \cite{pacmpl/ChenFHHHKLRW22,CHL+19}).
%Furthermore, numerous studies have demonstrated the broad applicability of SFTs across diverse domains. The work of \cite{VeanesHLMB12Transducer} showcases SFTs in security-critical applications, particularly for cross-site scripting (XSS) prevention. \cite{uss/HooimeijerLMSV11} extends this security focus by applying SFTs to web application sanitizer analysis. In system verification, \cite{osdi/YaseenABCL20} employs SFTs for runtime behavior monitoring, while \cite{pldi/HuD17} demonstrates their utility in automated program transformation through systematic inversion techniques.
%This symbolic approach not only provides a more succinct representation but also supports infinite alphabets, thereby extending the expressive capabilities of FAs and FTs.

@@

In summary, our formalization makes the following contributions:
\begin{enumerate}
\item We present the first formalization of symbolic finite transducers (SFTs) in the Isabelle/HOL. 
We prove the correctness of key operations between SFTs and symbolic SFAs, such as the product operation between SFTs and SFAs.
\item We leverage the refinement framework to create an extensible and efficient SFT formalization that is capable of accommodating various boolean algebras, ensuring adaptability to future transition label developments. The correctness of the refined efficient implementation is proved .
\item We develop a certified interval algebra with efficient set-theoretic operations, enabling refinement of transition labels from abstract sets to concrete intervals.
\item We demonstrate the practical utility of our formalization by applying it to sanitizer analysis and integrating our formalized SFT into the verified forward-propagation string solver CertiStr \cite{cpp/KanLRS22}. This integration extends CertiStr from supporting only simple concatenation and regular constraints to handling more sophisticated string replacement operations.
\end{enumerate}

>>>>

Symbolic Finite Automata (SFA) and Symbolic Finite Transducers (SFT)
\cite{cav/DAntoniV17, VeanesHLMB12Transducer,DV21} were proposed around 15 years ago
to deal with the alphabet-explosion problem that often arises in practical
applications. 
%represent advanced extensions of classic FAs and  FTs, enhancing their applicability in practical scenarios.
%
These symbolic extensions permit transition labels defined by an effective 
boolean algebra, allowing for a succinct representation of a set of transitions
that can be conveniently manipulated. 
%more expressive representations. 
For example, in the interval algebra, a transition label might be specified as 
an interval [$\texttt{a}-\texttt{z}$], encompassing all characters from
$\texttt{a}$ to $\texttt{z}$. As another example, in the SMT-Algebra
(particularly, of the theory of Linear Integer Arithmetic), %or 
one could associate an arithmetic condition ($x \mod 2 = 0$) to a transition, 
denoting all even numbers. Efficient algorithms for manipulating (e.g. computing
closures) and analyzing (e.g. checking emptiness) SFA and SFT are by now
rather mature (e.g see the CACM article \cite{DV21}).
SFA and SFT have been applied in numerous domains including analysis of web
applications against XSS vulnerabilities \cite{VeanesHLMB12Transducer,uss/HooimeijerLMSV11},
runtime behavior monitoring \cite{osdi/YaseenABCL20}, automated program
transformation \cite{pldi/HuD17}, and finally SMT solving for the unicode theory
of strings (e.g. see \cite{pacmpl/ChenFHHHKLRW22,CHL+19}).
%Furthermore, numerous studies have demonstrated the broad applicability of SFTs across diverse domains. The work of \cite{VeanesHLMB12Transducer} showcases SFTs in security-critical applications, particularly for cross-site scripting (XSS) prevention. \cite{uss/HooimeijerLMSV11} extends this security focus by applying SFTs to web application sanitizer analysis. In system verification, \cite{osdi/YaseenABCL20} employs SFTs for runtime behavior monitoring, while \cite{pldi/HuD17} demonstrates their utility in automated program transformation through systematic inversion techniques.
%This symbolic approach not only provides a more succinct representation but also supports infinite alphabets, thereby extending the expressive capabilities of FAs and FTs.

@@

In summary, our formalization makes the following contributions:
\begin{enumerate}
\item We present the first formalization of symbolic finite transducers (SFTs) in Isabelle/HOL.
We prove the correctness of key operations between SFTs and symbolic SFAs, such as the product operation between SFTs and SFAs.
\item We leverage the refinement framework to develop an extensible and efficient SFT formalization capable of supporting diverse Boolean algebras, ensuring adaptability to future transition label representations. The correctness of the refined, efficient implementation is proved via structural isomorphism between refinement layers.
\item We develop a certified interval algebra with verified set-theoretic operations, enabling refinement of transition labels from abstract sets to concrete intervals.
\item We demonstrate the practical utility of our formalization by applying it to sanitizer analysis and integrating our formalized SFT into the verified forward-propagation string solver CertiStr \cite{cpp/KanLRS22}. This integration extends CertiStr from supporting only simple concatenation and regular constraints to handling more sophisticated string replacement operations.
\end{enumerate}
